### 2021-03-23 assertions/doc/thesis-bernhardt/chapters/1-introduction.tex
Introduction revisited; offline design accuracy analysis

\chapter{Introduction}\label{chap:introduction}
Each execution of a blockchain-based Smart Contract in a gas-instrumented virtual machine generates costs proportional to the amount and complexity of computations and the  memory consumption \cite{goodman_tezos_2014}\cite{wood_ethereum_2021}. These fees, often calculated in gas, have to be paid for by the sender of the transaction and thus provides a powerful incentive to keep  smart contracts simple. However, in reality meaningful applications might need to perform some resource intensive computations at some point, which would be too expensive to execute inside the blockchain's virtual machine. Moreover, some blockchains impose a hard cap of how much gas can be consumed in one transaction, limiting the amount of operations that can be executed \cite{goodman_tezos_2014}.\\
In order to avoid these costs and restrictions, an application could perform some computations outside of the blockchain context (i.e., off-chain) and only submit the results to a smart contract as input parameter for further processing or to store it on the chain. This approach raises another problem though - trusting that the submitted results are correct poses a safety as well as a security risk. Invalid results can be submitted unintentionally, if the computation is faulty, or a third party may pass them intentionally to cause harm or take advantage. Consequently, submitted parameters need to be checked for specified properties before the smart contract is executed. Although in many cases, verification is less expensive than computing the result itself, the generated costs for checking the validity increase with the size of the problem or input, ultimately becoming too expensive. \\
Consider, for instance, the following contract from \cite{thiemann_2020} that expects a prime number: \todo{State programming language?!}
\begin{lstlisting}[caption=Smart contract expecting a prime number, numbers=none, language=Solidity, label=lst:prime]
contract Example {
	function ( int p ) public {
	// assume p is prime
	...
	}
}
\end{lstlisting}

The assumption that the given integer $p$ is a prime can be expressed as an explicit assertion in predicate logic: 
\begin{equation}\label{eq:prime}
    (\forall n) (2 \le n \le \sqrt{p}) \Rightarrow (p \% n) \neq 0
\end{equation}

Checking if this property holds requires the contract to execute a loop in $\mathcal{O}(\sqrt{p})$ time and therefore generate costs linear in $\sqrt{p}$ on top of the actual contract execution, assuming computing the body of the loop takes constant time.\\
Thus it appears that we need a mechanism that allows the smart contract to make use of the computed result from an off-chain source without having to validate its validity locally with scaling costs.

\section{A distributed approach for assertion checking}
A more efficient way to solve the issue at hand is to engage the validators of the blockchain into a distributed effort to find a counterexample that disproves an assumption. If no counterexample could be found by any validator, the assumption is considered to be valid and the execution of the contract may resume. On the other hand, an identified counterexample is made public on the peer-to-peer network and thus cancels the contract execution (i.e., the transaction fails). \\
In order to find counterexamples of an assertion, the formula in \ref{eq:prime} has to be transformed to its negation:
\begin{equation}\label{eq:prime_neg}
	(\exists n) (2 \le n \le \sqrt{p}) \wedge (p \% n) = 0
\end{equation}

>>>>

\chapter{Introduction}\label{chap:introduction}
Headed by Bitcoin in 2009, ``cryptocurrencies have emerged as the first generation of blockchain technology'' \cite{alharby_blockchain_2017}. However, their transactions are mostly limited to simple monetary transactions or registration of assets or ownerships \cite{alharby_blockchain_2017}. More recent blockchains, considered as the second generation \cite{alharby_blockchain_2017}, aim to provide a platform for building more sophisticated and complex applications on top of the technology. These are built with smart contracts, programs written in Turing-complete languages. Running a smart contract, though, generates costs proportional to the amount and complexity of computations, as well as its storage usage \cite{chen_under-alharby_blockchain_2017}. These fees, often measured in a unit called ``gas'', have to be paid for by the user in order to compensate the miners for their efforts \cite{chen_under-optimized_2020}. This creates a powerful incentive to keep smart contracts simple and efficient, while in effect, meaningful applications still might need to perform some resource intensive computations. Not only can such applications generate unsustainable costs when executed on the blockchain, but they may also reach gas limits imposed by the blockchain's protocol. \\
In order to avoid these costs and restrictions, such an application could perform some computations outside of the blockchain context (i.e., off-chain) and only submit the results to a smart contract as an input parameter for further processing or recording \cite{thiemann_2020}. By trusting that the submitted results are correct, this approach raises both safety and security issues. Invalid results can be submitted unintentionally as a result of a wrong computation, or a third party may pass them intentionally to cause harm or take advantage. Consequently, submitted parameters should be checked for assumed properties before the smart contract is executed. Although in many cases, verification is less expensive than computing the result itself\footnote{cf. sorting algorithms with average time complexities of $\mathcal{O}(n log n)$ - $\mathcal{O}(n^2)$ \cite{review_sorting} vs. linear time to check if it is sorted.}, the generated costs for checking the validity increase with the size of the input. \\
Consider, for instance, the following contract:
\begin{lstlisting}[caption=Smart contract expecting a prime number \cite{thiemann_2020}, numbers=none, language=Solidity, label=lst:prime]
contract Example {
	function ( int p ) public {
	// assume p is prime
	...
	}
}
\end{lstlisting}
The assumption, that the given integer $p$ is a prime, can be expressed as an explicit assertion in predicate logic \cite{thiemann_2020}: 
\begin{equation}\label{eq:prime}
    (\forall n) (2 \le n \le \sqrt{p}) \Rightarrow (p \% n) \neq 0
\end{equation}
To check whether this property holds for $p$, the contract has to execute a loop. Assuming computing the body of the loop takes constant time, this test has a time complexity of $\mathcal{O}(\sqrt{p})$ and therefore generates costs linear in $\sqrt{p}$ on top of the cost for executing the actual contract. Thus, a mechanism is required that allows the smart contract to make use of the result computed by an off-chain source, without having to check its validity locally with scaling costs.

\section{A distributed approach for assertion checking}
% analyse/propose/explore an approach that could solve this rather than "this is the solution"
A more efficient way to solve the issue at hand is to engage the validators of the blockchain into a distributed effort to find a counterexample that disproves an assumption. If no counterexample could be found by any validator, the assumption is considered to be valid and the execution of the contract may resume. On the other hand, an identified counterexample is made public on the peer-to-peer network and thus cancels the contract execution (i.e., the transaction fails). \\
In order to find counterexamples of an assertion, the formula in \ref{eq:prime} has to be transformed to its negation:
\begin{equation}\label{eq:prime_neg}
	(\exists n) (2 \le n \le \sqrt{p}) \wedge (p \% n) = 0
\end{equation}

### 2021-03-23 assertions/doc/thesis-bernhardt/chapters/2-offline_design.tex
Introduction revisited; offline design accuracy analysis

\section{Accuracy of the approach}
Since the proposed approach implements a probabilistic test, its accuracy as well as the costs incurred by guaranteeing a specified certainty threshold, remains to be analysed. The goal is to identify a formula that, given the domain of a formula in predicate logic, returns an estimate of a lower bound of samples necessary to find an existing counterexample with a probability $p$. From this analysis, it can be derived how effective a formula can be checked on a blockchain with $m$ validators. For the remainder of this chapter, the assumptions hold that the random generator picks elements independently from a uniform distribution and that there exists exactly one counterexample for the passed parameter.
\todo{further measures for blockchains with small number of validators}

@@

\subsection{Setting probability thresholds}

effectiveness:
smaller search spaces = less needed test runs 
example sorted list -> generate 1 random number and check neighbor = many test runs to cover all of them
partition search space \& generate 2 random numbers better? -> 2 smaller search spaces. would this improve anything?

\begin{lstlisting}[numbers=none, language=Assertion]
(entrypoint (left (right (pair ( a : address) _)))
  (forall (i : int)
     (if (gt i 10)
       (assert true))))
\end{lstlisting}

>>>>

\section{Accuracy of the approach}
Since the proposed approach implements a probabilistic test, its accuracy as well as the costs incurred by guaranteeing a specified certainty threshold, remains to be analysed. The goal is to identify a formula that, given the domain of a formula in predicate logic, returns an estimate of a lower bound of samples necessary to find an existing counterexample with a probability $p$. From this analysis, it can be derived how effective a formula can be checked on a blockchain with $m$ validators. For the remainder of this chapter, the assumptions hold that the random generator picks elements independently from a uniform distribution and that there exists exactly one counterexample for the passed parameter.\\
The following two sections depict the findings from \cite{bernhardt_veigel_2020}.
\todo{further measures for blockchains with small number of validators}
\todo{Refer to our example paper}

@@

\subsection{Setting probability thresholds}
The goal is to find a number $t$ of test runs, s.t. the probability $P_{c,t}$ of not finding the counterexample drops below a certain threshold $c$. A validator finds the counterexample in one test run with probability $\frac{1}{n}$, and misses it with probability $(1-\frac{1}{n})$. After $t$ tests runs, the probability that the counterexample has not been found is thus $(1-\frac{1}{n})^t$. Following the same approach as in \cite{mahl_schindel_2007} to retrieve a bound for $P_{c,t}$, we use
\begin{equation}
(1-\frac{1}{m})^m \leq \frac{1}{e}
\end{equation}
which holds for all $m > 0$. From this inequality, it follows that
\begin{equation}
(1-\frac{1}{n})^t = ((1-\frac{1}{n})^n)^{\frac{t}{n}} \le e^{-\frac{t}{n}}
\end{equation}
With this, the probability $P_{c,t}$ of not finding the counterexample with $t$ test runs can be defined as
\begin{equation}
P_{c,t} \le e^{-\frac{t}{n}}
\end{equation}
In order to retrieve the number of test runs $t$ necessary for $P_{c,t}$ to drop below a specified threshold $c$, the inequality is solved for $t$, which depends on the known parameter $n$ and an arbitrary value for $c$:
\begin{align}
    e^{-\frac{t}{n}} &\leq c && \text{with } c\le1 \nonumber\\
    t &\geq -n\:\ln(c)
\end{align}
For $c = \frac{1}{e}$ ($\approx 36.79\:\%$) $t$ is bound by $n$, which means that for higher reliabilities the number of test runs needs to be greater than $n$. \figref{fig:graph_t_c} shows the lower bounds of $t$ as a function of the probability threshold and domain size. One can see that the increase in accuracy is approximately linear to the number of test runs below a threshold of $c=0.5$, but requires an exponential testing effort to reach higher thresholds.
\begin{figure}[h]
\includegraphics[width=0.95\linewidth]{figures/3-offline/graph_t_c}
\label{fig:graph_t_c}
\end{figure}
For assertions with several quantifiers, which are translated to random generators, all domains have to be considered in the calculation. Revisiting the intersection of sets from \secref{sec:existential}, every element in set $u$ has to be compared to every element in set $v$. Thus, the lower bound of test runs needed to check they intersect with a certainty of 63.21\% is given by $t \leq t_1 * t_2$ with $t_1 \leq |u|$ and $t_2 \leq |v|$.

\subsection{Validators vs. test runs}
The last section derived a formula to calculate a lower bound for a number of test runs $t$ in order to reach a certain level of confidence. However, $t$ is not a parameter that can be adjusted to an arbitrary value on the blockchain, as the assertion is checked by the validators. Assuming a blockchain has $m$ validators, the actual lower bound of test runs is $m$. For cases where there are not enough validators to reach a threshold $c$, a mechanism is required to have each validator to check the property for multiple random values. Hence, the actually executed number of checks is always a multiple of $m$.

\subsection{Alternatives to random testing}
For applications where higher reliability is crucial, an alternative to checking elements randomly could be to coordinate a systematic iteration of the domain. The implementation of such a coordination is not trivial though; validators can't simply be passed an individual value as input. Instead, possible approaches could be to implement a central instance that allocates an element in the search space to each validator, or use some unique and intrinsic attribute, like an identification number, as input for a mapping.

\subsubsection{Central instance assigning values}
For this approach, the tool-chain would need to generate a separate contract that is called by the baker and validators as a proxy. This contract would then, for instance, generate a number using a modulo-n counter and pass this number as an additional parameter to the assertion code. The random generators would be omitted accordingly. However, this works only as expected if there are no simultaneous transactions calling the contract, otherwise it cannot be guaranteed that all elements have been assigned for each transaction. If and how this issue can be solved will not be discussed further in this thesis.

\subsubsection{Using unique attributes of a validator}\label{sec:alt_attributes}
This approach strongly depends on whether the validators have a unique id or other attribute that can be accessed from within a contract.  If there is (or the language can be extended with such a feature), there needs to exist a non-injective surjective function that maps the respective attributes represented by type A to an element of the domain represented by type B. Furthermore, for cases where $t > m$, an offset needs to be added, s.t. a validator checks distinct elements for each test run. \\
Implementing such an approach on Tezos does become even more challenging by the way endorsing rights are distributed in its proof-of-stake mechanism: for each block level, endorsing rights are assigned to the owner of a randomly selected roll, i.e., a set of tokens \cite{tezos_docs}. This means that the same endorser can be picked multiple times for endorsing one block and thus some elements in the domain may remain unchecked. \secref{lala} goes into more detail about this issue. \todo{Ref section}

### 2021-03-23 assertions/doc/thesis-bernhardt/chapters/2-tezos.tex
Offline Chapter beginnign
Use Case chapter
Introduction revisited; offline design accuracy analysis

\chapter{Tezos}
The Tezos blockchain is presented in its Whitepaper as a \enquote{generic and self-amending crypto-ledger} \cite{goodman_tezos_2014}. It uses a proof-of-stake consensus mechanism that is not only used to agree on the current state of its ledger, but also allows its stakeholders to come to a consensus about changes in the economic protocol by a voting process. These changes included in a protocol upgrade, i.e. amendment, can influence which transactions are valid on the blockchain, the payment system or even the voting process itself without risking a fork of the blockchain. Everyone owning the cryptocurrency of Tezos, called Tez, is considered a stakeholder. The Whitepaper compares the self-amending protocol to Peter Suber's game Nomic \cite{nomic}, and a similar concept was also implemented in the virtual sports league Blaseball \cite{blaseball}, which became popular during the COVID-19 pandemic. In addition to user accounts associated with a public key, Tezos supports smart contracts, which are written in the built-in language Michelson. Tezos' transaction fee system is similar to Ethereum's - it is gas-instrumented and besides a base fee, every operation and byte of storage has to be paid for. However, Tezos imposes a hard cap on the amount of gas that can be consumed per transaction, whereas Ethereum limits the gas quota only in respect to blocks \cite{wood_ethereum_2021}.\\
Compiling the Tezos source code yields five essential binaries: the node, baker, endorser, accuser and client. The node is the entity that connects to the peer-to-peer network and keeps a copy of the chain. Bakers are responsible for producing new blocks, the endorsers for signing blocks and the accusers to call out bakers or endorsers which double-sign or -endorse. The client provides a command line interface to interact with the local node through remote procedure calls (RPC).

\section{Michelson}
Michelson is a lower-level, stack-based language with a strict type-checking and supports primitive data types, like integers or strings, as well as high-level data structures such as list, maps and sum types \cite{tezos_docs_michelson}. The type system reduces the occurrence of runtime errors and ensures that only well-typed contracts are originated on the blockchain. In OCaml the typing rules are implemented using generalized algebraic data types (GADTs) \cite{tezos_docs_michelson} and thus enforced by OCaml's typechecker. \\
The concrete syntax of Michelson is called Micheline. A program is represented in Micheline nodes, which can be a constant of type integer, string or byte sequence, or an application of a primitive, given as a string keyword, to a list of nodes. For documentation, readability and additional type constraints, Michelson and Micheline also offer three types of annotations - type, variable and field or constructor annotations, which are labelled with a unique special character in Micheline. The toplevel structure of a smart contract consists of a sequence of the three primitive applications \texttt{parameter, storage} and \texttt{code}, declaring the type of the input parameter, the storage type and the actual program code. Unlike in the contracts given in the introduction, described with the syntax of Solidity, Michelson doesn't have a concept of named functions with individual input parameter types.

@@

\begin{itemize}
\item[Flexible Network Sandboxes (Flextesa)] With Flextesa one can configure and run a small, fully functional sandboxed test network including nodes, bakers, endorsers and accusers. Accounts can be instantiated and used to send or receive funds. Configurable are, for instance, the size of the network, the time between blocks or enabling generation of random network traffic. Networks may either be fully autonomous, baking blocks automatically, or require manual baking of blocks. It also supports interactive sessions, where the tester can interact with the blockchain using the Tezos client. Flextesa's typical use-cases are interactive testing scenarios like double-baking, the voting process or protocol amendments. The Tezos repository already provides ready-to-use scenarios for accusations. Besides testing shell or protocol code, Flextesa can also be used to test smart contracts.\cite{tezos docs}
\item[Tezt] The Tezt framework is newer than Flextesa and planned as its replacement. It can also launch the Tezos binaries as external processes and build a sandboxed test network of configurable size. Its main advantages over Flextesa are simplicity, usability and the use of events instead of polling the node.\cite{tezos_docs_tezt}
\end{itemize}

\subsection{High-level languages}
Michelson is a compilation target for various high-level languages that provide more user-friendly and intuitive way of writing smart contracts. Additionally, some of them come with development environments, testing or verification tools. The following list comprises of the three most prominent languages that compile to Michelson:

\begin{itemize}
\item[\textbf{Liquidity}] is a language with an Ocaml-like syntax, which allows using local variables instead of stack manipulations. Its module system can be used to write reusable contract code or libraries. Besides an optimizing compiler, the project also includes a decompiler to compile Michelson programs to Liquidity. \cite{liquidity}
\item[\textbf{SmartPy}] is a language available through a Python library and lets developers write contracts and tests in Python. Its developer suite includes i.a. a compiler, a simulation engine and an online editor. \cite{smartpy}
\item [\textbf{fi}] 's syntax similar to JavaScript or Ethereum's contract language Solidity. It also provides an online editor and a simulator to test against the compiled contracts. However, it seems extensions to Michelson introduced by newer protocol versions are not supported yet. \cite{https://fi-code.com/}
\end{itemize}

>>>>

\chapter{Tezos}
The Tezos blockchain is presented in its whitepaper as a \enquote{generic and self-amending crypto-ledger} \cite{goodman_tezos_2014}. It uses a proof-of-stake consensus mechanism that is not only used to agree on the current state of its ledger, but also allows its stakeholders to come to a consensus about changes in the economic protocol by a voting process. These changes included in a protocol upgrade, i.e. amendment, can influence which transactions are valid on the blockchain, the payment system or even the voting process itself without risking a fork of the blockchain. Everyone owning the cryptocurrency of Tezos, called Tez , is considered a stakeholder. The whitepaper compares the self-amending protocol to Peter Suber's game Nomic \cite{nomic}, and a similar concept was also implemented in the virtual sports league Blaseball \cite{blaseball}, which became popular during the COVID-19 pandemic. In addition to user accounts associated with a public key, Tezos supports smart contracts, which are written in the built-in language Michelson. Tezos' transaction fee system is similar to Ethereum's - it is gas-instrumented and besides a base fee, every operation and byte of storage has to be paid for. However, Tezos imposes a hard cap on the amount of gas that can be consumed per transaction, whereas Ethereum limits the gas quota only in respect to blocks \cite{wood_ethereum_2021}.\\
Compiling the Tezos source code yields five essential binaries: the node, baker, endorser, accuser and client. The node is the entity that connects to the peer-to-peer network and keeps a copy of the chain. Bakers are responsible for producing new blocks, the endorsers for signing blocks and the accusers to call out bakers or endorsers which double-sign or -endorse. The client provides a command line interface to interact with the local node through remote procedure calls (RPC).

\section{Proof-of-stake in Tezos}
In Tezos, contracts that staked a minimum amount of tokens, called a roll, can participate in the consensus mechanism and can either have the role of a baker or endorser. Contracts who don't own enough tokens or infrastructure to participate directly can delegate their baking and endorsing rights to other contracts. The rights are determined and assigned at the beginning of each cycle, which consists of a specified number of blocks. For baking, a random roll is selected for each block level and the rights are assigned to its owner. The block produced by that baker is endorsed by a fixed number of endorsers (currently 32 in protocol 008 Edo), which have been assigned endorsing rights for this block level if their roll has been randomly selected. Since participants can stake more than one roll, an endorser may be assigned several endorsement slots at the same level. \\
As an incentive for active participation in the consensus algorithm, delegates (and also delegators) receive rewards in form of tokens. However, if accusers detect double-baking or -endorsement, the delegate is penalized by burning (i.e., destroying) their security deposit. \cite{tezos_docs}

\section{Michelson}
Michelson is a lower-level, stack-based language with a strict type-checking and supports primitive data types, like integers or strings, as well as high-level data structures such as list, maps and sum types \cite{tezos_docs}. The type system reduces the occurrence of runtime errors and ensures that only well-typed contracts are originated on the blockchain. In OCaml the typing rules are implemented using generalized algebraic data types (GADTs) \cite{tezos_docs} and thus enforced by OCaml's typechecker. \\
The concrete syntax of Michelson is called Micheline. A program is represented in Micheline nodes, which can be a constant of type integer, string or byte sequence, or an application of a primitive, given as a string keyword, to a list of nodes. For documentation, readability and additional type constraints, Michelson and Micheline also offer three types of annotations - type, variable and field or constructor annotations, which are labelled with a unique special character in Micheline. The toplevel structure of a smart contract consists of a sequence of the three primitive applications \texttt{parameter, storage} and \texttt{code}, declaring the type of the input parameter, the storage type and the actual program code. Unlike in the contracts given in the introduction, described with the syntax of Solidity, Michelson doesn't have a concept of named functions with individual input parameter types.

\todo{Citations}
\todo{Put Michelsons grammar/list of operations in appendix?}

@@

\begin{itemize}
\item[Flexible Network Sandboxes (Flextesa)] With Flextesa one can configure and run a small, fully functional sandboxed test network including nodes, bakers, endorsers and accusers. Accounts can be instantiated and used to send or receive funds. Configurable are, for instance, the size of the network, the time between blocks or enabling generation of random network traffic. Networks may either be fully autonomous, baking blocks automatically, or require manual baking of blocks. It also supports interactive sessions, where the tester can interact with the blockchain using the Tezos client. Flextesa's typical use-cases are interactive testing scenarios like double-baking, the voting process or protocol amendments. The Tezos repository already provides ready-to-use scenarios for accusations. Besides testing shell or protocol code, Flextesa can also be used to test smart contracts.\cite{tezos_docs}
\item[Tezt] The Tezt framework is newer than Flextesa and planned as its replacement. It can also launch the Tezos binaries as external processes and build a sandboxed test network of configurable size. Its main advantages over Flextesa are simplicity, usability and the use of events instead of polling the node.\cite{tezos_docs}
\end{itemize}

\subsection{High-level languages}
Michelson is a compilation target for various high-level languages that provide more user-friendly and intuitive way of writing smart contracts. Additionally, some of them come with development environments, testing or verification tools. The following list comprises of the three most prominent languages that compile to Michelson:
\begin{itemize}
\item[\textbf{Liquidity}] is a language with an Ocaml-like syntax, which allows using local variables instead of stack manipulations. Its module system can be used to write reusable contract code or libraries. Besides an optimizing compiler, the project also includes a decompiler to compile Michelson programs to Liquidity. \cite{liquidity}
\item[\textbf{SmartPy}] is a language available through a Python library and lets developers write contracts and tests in Python. Its developer suite includes i.a. a compiler, a simulation engine and an online editor. \cite{smartpy}
\item [\textbf{fi}] 's syntax similar to JavaScript or Ethereum's contract language Solidity. It also provides an online editor and a simulator to test against the compiled contracts. However, it seems extensions to Michelson introduced by newer protocol versions are not supported yet. \cite{https://fi-code.com/}
\end{itemize}

### 2021-03-23 assertions/doc/thesis-bernhardt/chapters/2-use_cases.tex
Started rework of Offline design
Introduction revisited; offline design accuracy analysis

\section{Definition of a set of logical formulae}\label{sec:formulae}
\begin{align*}
    t &::= \text{primitive data types (existing in the blockchain VM)} \\
    \sigma &::= \forall \mid \exists \\
    \Theta &::= \Phi \mid \sigma (x:t) \Theta \\
    \Phi,\Psi &::= \neg\Phi \mid \Phi \Rightarrow \Psi \mid \Phi\wedge\Psi \mid
    				\Phi\vee\Psi \mid \Phi \oplus \Psi \mid M \rho N \\
    \rho &::= < \mid > \mid \le \mid \ge \mid = \mid \ne \\
    M, N &::= x \mid c \mid M \odot N  \mid f (\overline M) \\
    \odot &::= +\mid -\mid * \mid / \mid \% \\
    c &::= \text{constants: numbers, strings, } \top, \bot \\
    f &::= \text{operations (existing in the blockchain VM)}
\end{align*}
\begingroup\vspace*{-\baselineskip}
\captionof{figure}{Set of logical formulas amenable to distributed assertion checking (extension of set given in \cite{thiemann_2020})}
\vspace*{\baselineskip}\endgroup

@@

\section{Existential quantifier and proofs of validity}
The examples given so far in this thesis used universal quantification and required a search for counterexamples in order to check if some assumption about an input parameter is true. However, the same cannot be applied to existential quantifiers. Consider a contract expecting a string $e$ and a list $l$ of strings that contains $e$ at least once. The assumption that \texttt{l} contains \texttt{e} can be expressed by
\begin{equation}\label{eq:contains}
	(\exists n : int) (0 \leq n < |l|) \Rightarrow l[n] = e
\end{equation}
and its negation is 
\begin{equation}\label{eq:contains_neg}
	(\forall n : int) (0 \leq n < |l|) \Rightarrow l[n] \ne e
\end{equation}

If the assertion fails for a random value of $n$, it does not produce a counterexample, but rather it produces a proof for the validity of the assumption. In that case, the validator finding such a proof should publish an approval instead of a veto to the network and the normal execution of the contract can proceed. Hence, there must be two different modi of assertion checking that implement different
types of assertions (proofs need assertions that don't abort the program).
\todo{Refer to a chapter where this is discussed further?}

@@

\subsection{Further nesting}
Nesting is, of course, not limited to two quantifiers, but can contain many of them. Generally, formulas containing no quantification alterations benefit the most from distributed assertion checking, as each quantifier can be translated into a random generator. On the other hand, formulas with a quantifier alteration early in the order of quantifiers benefit the least. Starting from the first alteration, the quantifiers are translated to loops.\\
To substantiate this argument, consider a two variations of a contract that expect a list of boolean formulas in DNF, and an assignment $\mathcal{A}$. The first one checks, whether $\mathcal{A}$ satisfies at least one $F$, the other one checks if all $F$ are satisfied. This can be expressed by nesting quantifications as follows: 
\begin{itemize}
\item $\exists f \exists i \forall j$
\item $\forall f \exists i \forall j$
\end{itemize}
In the first case, each validator can pick a random clause in a random $F$ and check, if all literals evaluate to true. In the second case, the validators have to find a counterexample, namely a clause that is not satisfied by \mathcal{A}. Checking a random clause in isolation cannot disprove the satisfiability, hence the validator has to check all clauses.

@@

\section{Cost analysis} \todo{this}
%diagram cost increase?
%experiment: write Liquidity contract w. 1 loop and 1 single execution * loop size

>>>>

\section{Definition of a set of logical formulae}\label{sec:formulae}
\begin{align*}
    t &::= \text{primitive data types (existing in the blockchain VM)} \\
    \sigma &::= \forall \mid \exists \\
    \Theta &::= \Phi \mid \sigma (x:t) \Theta \\
    \Phi,\Psi &::= \neg\Phi \mid \Phi \Rightarrow \Psi \mid \Phi\wedge\Psi \mid
    				\Phi\vee\Psi \mid \Phi \oplus \Psi \mid M \rho N \\
    \rho &::= < \mid > \mid \le \mid \ge \mid = \mid \ne \\
    M, N &::= x \mid c \mid M \odot N  \mid f (\overline M) \\
    \odot &::= +\mid -\mid * \mid / \mid \% \\
    c &::= \text{constants: numbers, strings, } \top, \bot \\
    f &::= \text{operations (existing in the blockchain VM)}
\end{align*}
\begingroup\vspace*{-\baselineskip}
\captionof{figure}{Set of logical formulas amenable to distributed assertion checking (extension of set from \cite{thiemann_2020})}
\vspace*{\baselineskip}\endgroup

@@

\section{Existential quantifier and proofs of validity}\label{sec:existential}
The examples given so far in this thesis used universal quantification and required a search for counterexamples in order to check if some assumption about an input parameter is true. However, the same cannot be applied to existential quantifiers. Consider a contract expecting two intersecting sets of integers, represented as arrays. This property can be checked in $\mathcal{O}(|u|*|v|)$. With $n = |u|+|v|$, the worst case for the run time occurs when both sets have the same size, i.e., $|u| = |v| = n/2$. This leads to a quadratic complexity $\mathcal{O}(n/2 * n/2) = \mathcal{O}(n^2/4) = \mathcal{O}(n^2)$.\\
The assumption, that the intersection of the sets is not empty, i.e., $u \cap v \neq \emptyset$, can be expressed by
\begin{equation}\label{eq:intersect}
  (\exists i, \exists j) (0 \le i < |u|) \land (0 \le j < |v|) \Rightarrow u[i] = v[j]
\end{equation}
and its negation is
\begin{equation}\label{eq:intersect_neg}
  (\forall i, \forall j) (0 \le i < |u|) \land (0 \le j < |v|) \land u[i] \neq v[j]
\end{equation}
If the assertion `fails' for random values of $i,j$, it does not produce a counterexample, but it rather produces a proof for the validity of the assumption. In that case, the validator finding such a proof should publish an approval instead of a veto to the network and the normal execution of the contract can proceed. Hence, there must be two different modi of assertion checking that implement different types of assertions.
\todo{Refer to a chapter where this is discussed further?}

@@

\subsection{Further nesting}
Nesting is, of course, not limited to two quantifiers, but can contain many of them. Generally, formulas containing no quantification alterations benefit the most from distributed assertion checking, as each quantifier can be translated into a random generator. On the other hand, formulas with a quantifier alteration early in the order of quantifiers benefit the least. Starting from the first alteration, the quantifiers are translated to loops.\\
To substantiate this argument, consider a two variations of a contract that expect a list of boolean formulas in DNF, and an assignment $\mathcal{A}$. The first one checks, whether $\mathcal{A}$ satisfies at least one $F$, the other one checks if all $F$ are satisfied. This can be expressed by nesting quantifications as follows: 
\begin{itemize}
\item $\exists f \exists i \forall j$
\item $\forall f \exists i \forall j$
\end{itemize}
In the first case, each validator can pick a random clause in a random $F$ and check, if all literals evaluate to true. In the second case, the validators have to find a counterexample, namely a clause that is not satisfied by $\mathcal{A}$. Checking a random clause in isolation cannot disprove the satisfiability, hence the validator has to check all clauses.

@@

\section{Cost analysis} \todo{this}
% diagram cost increase?
% experiment: write Liquidity contract w. 1 loop and 1 single execution * loop size
